<link rel="stylesheet" href="http://netdna.bootstrapcdn.com/bootstrap/3.0.3/css/bootstrap.min.css">
<style type="text/css"> body {padding: 10px 30px 10px 30px;} table,th, td {text-align: center;} </style>


Case Solution - Factor Analysis 
========================================================

**Jason Ng
Nihar R Prasad
Jesus Martin
Priya Sankarlingam
Jakob Rieckh**

Introduction
---------------------------------------------------------

We have used Boats file with questions 1 to 29. These questions were essentially asked from different customers about their experiences with boats.

When we started doing cluster analysis, we got a club of five major factors. The factoring analysis essentially resulted in five major factors with eigen value greater than 1 and cumulative percentage of explained variance 52.74. So, we decided to club the factors with labels with commonality.

The five factors were:
Component 1(Qs 9,26,3,13,27,28,14,16,7,17,20): Image/Status Symbol
Component 2(Qs 18,21,19,24,6,22,7): Leisure
Component 3(Qs 4,5,15):Brand awareness/high-end preference
Component 4(Qs 11,16,17): Boat Expertise
Component 5(Qs 2, 12, 8): Low-end preference

However, since high-end and low-end prefernces both relate to price sensitivity, we collapsed the two factors into price sensitivity.


The Data
--------------------------------------------


```{r echo=FALSE, message=FALSE, prompt=FALSE, results='asis'}
# let's make the data into data.matrix classes so that we can easier visualize them
ProjectDataFactor = data.matrix(ProjectDataFactor)
ProjectData = data.matrix(ProjectData)
```

<div class="row">
<div class="col-md-6">
```{r echo=FALSE, message=FALSE, prompt=FALSE, results='asis'}
show_data = data.frame(round(ProjectDataFactor,2))
show_data = show_data[1:min(max_data_report,nrow(show_data)),]
row<-rownames(show_data)
dfnew<-cbind(row,show_data)
change<-colnames(dfnew)
change[1]<-"Variables"
colnames (dfnew)<-change
m1<-gvisTable(dfnew,options=list(showRowNumber=TRUE,width=1220, height=min(400,27*(nrow(show_data)+1)),allowHTML=TRUE,page='disable'))
print(m1,'chart')
```
</div>
</div>

#### Factor Analysis in 6 steps

1. Confirm the data in metric 

2. Decide whether to scale or standardize the data

3. Check the correlation matrix to see if Factor Analysis makes sense

4. Develop a scree plot and decide on the number of factors to be derived

5. Interpret the factors (consider factor rotations - technical but useful)

6. Save factor scores for subsequent analyses

Let's follow these steps.

#### Step 1: Confirm the data are metric


<div class="row">
<div class="col-md-6">
```{r echo=FALSE, comment=NA, warning=FALSE, message=FALSE,results='asis'}
show_data = data.frame(round(my_summary(ProjectDataFactor),2))
#show_data = show_data[1:min(max_data_report,nrow(show_data)),]
row<-rownames(show_data)
dfnew<-cbind(row,show_data)
change<-colnames(dfnew)
change[1]<-"Variables"
colnames (dfnew)<-change
m1<-gvisTable(dfnew,options=list(showRowNumber=TRUE,width=1220, height=min(400,27*(nrow(show_data)+1)),allowHTML=TRUE,page='disable'))
print(m1,'chart')
```
</div>
</div>


#### Step 2: Decide whether to scale or standardize the data


```{r, results='asis'}
ProjectDatafactor_scaled=apply(ProjectDataFactor,2, function(r) {if (sd(r)!=0) res=(r-mean(r))/sd(r) else res=0*r; res})
```


<div class="row">
<div class="col-md-6">
```{r echo=FALSE, comment=NA, warning=FALSE, message=FALSE,results='asis'}

show_data = data.frame(round(my_summary(ProjectDatafactor_scaled),2))
#show_data = show_data[1:min(max_data_report,nrow(show_data)),]
row<-rownames(show_data)
dfnew<-cbind(row,show_data)
change<-colnames(dfnew)
change[1]<-"Variables"
colnames (dfnew)<-change
m1<-gvisTable(dfnew,options=list(showRowNumber=TRUE,width=1220, height=min(400,27*(nrow(show_data)+1)),allowHTML=TRUE,page='disable'))
print(m1,'chart')
```
</div>
</div>


#### Step 3:  Check correlation matrix to see if Factor Analysis makes sense


```{r echo=FALSE, comment=NA, warning=FALSE, message=FALSE, results='asis'}
thecor = round(cor(ProjectDataFactor),2)
colnames(thecor)<-colnames(ProjectDataFactor)
rownames(thecor)<-colnames(ProjectDataFactor)
## printing the result in a clean-slate table
#cat(renderHeatmapX(thecor, border=1,center = 0,vrange_up = 1, vrange_down = 1))
cat(renderHeatmapX(thecor, border=1))
    
```


#### Step 4. Develop a scree plot and decide on the number of factors to be derived


```{r echo=FALSE, comment=NA, warning=FALSE, message=FALSE,results='asis'}
UnRotated_Results<-principal(ProjectDataFactor, nfactors=ncol(ProjectDataFactor), rotate="none",score=TRUE)
UnRotated_Factors<-round(UnRotated_Results$loadings,2)
UnRotated_Factors<-as.data.frame(unclass(UnRotated_Factors))
colnames(UnRotated_Factors)<-paste("Component",1:ncol(UnRotated_Factors),sep=" ")
```


```{r echo=FALSE, comment=NA, warning=FALSE, error=FALSE,message=FALSE,results='asis'}
Variance_Explained_Table_results<-PCA(ProjectDataFactor, graph=FALSE)
Variance_Explained_Table<-Variance_Explained_Table_results$eig
Variance_Explained_Table_copy<-Variance_Explained_Table


row=1:nrow(Variance_Explained_Table)
name<-paste("Component No:",row,sep="")
Variance_Explained_Table<-cbind(name,Variance_Explained_Table)
Variance_Explained_Table<-as.data.frame(Variance_Explained_Table)
colnames(Variance_Explained_Table)<-c("Components", "Eigenvalue", "Percentage_of_explained_variance", "Cumulative_percentage_of_explained_variance")

m<-gvisTable(Variance_Explained_Table,options=list(width=1200, height=min(400,27*(nrow(show_data)+1)),allowHTML=TRUE,page='disable'),formats=list(Eigenvalue="#.##",Percentage_of_explained_variance="#.##",Cumulative_percentage_of_explained_variance="#.##"))
print(m,'chart')
```


```{r Fig1, echo=FALSE, comment=NA, results='asis', message=FALSE, fig.align='center', fig=TRUE}

eigenvalues  <- Variance_Explained_Table[,2]
df           <- cbind(as.data.frame(eigenvalues), c(1:length(eigenvalues)), rep(1, length(eigenvalues)))
colnames(df) <- c("eigenvalues", "components", "abline")
Line         <- gvisLineChart(as.data.frame(df), xvar="components", yvar=c("eigenvalues","abline"), options=list(title='Scree plot', legend="right", width=900, height=600, hAxis="{title:'Number of Components', titleTextStyle:{color:'black'}}", vAxes="[{title:'Eigenvalues'}]",  series="[{color:'green',pointSize:3, targetAxisIndex: 0}]"))
print(Line, 'chart')
```


```{r echo=FALSE, comment=NA, warning=FALSE,message=FALSE,results='asis'}
if (factor_selectionciterion == "eigenvalue")
  factors_selected = sum(Variance_Explained_Table_copy[,1] >= 1)
if (factor_selectionciterion == "variance")
  factors_selected = 1:head(which(Variance_Explained_Table_copy[,"cumulative percentage of variance"]>= minimum_variance_explained),1)
if (factor_selectionciterion == "manual")
  factors_selected = manual_numb_factors_used

```


#### 5. Interpret the factors (consider factor rotations - technical but useful)


```{r echo=FALSE, comment=NA, warning=FALSE, message=FALSE,results='asis'}

Rotated_Results<-principal(ProjectDataFactor, nfactors=max(factors_selected), rotate=rotation_used,score=TRUE)
Rotated_Factors<-round(Rotated_Results$loadings,2)
Rotated_Factors<-as.data.frame(unclass(Rotated_Factors))
colnames(Rotated_Factors)<-paste("Component",1:ncol(Rotated_Factors),sep=" ")

sorted_rows <- sort(Rotated_Factors[,1], decreasing = TRUE, index.return = TRUE)$ix
Rotated_Factors <- Rotated_Factors[sorted_rows,]

show_data <- Rotated_Factors 
#show_data = show_data[1:min(max_data_report,nrow(show_data)),]
row<-rownames(show_data)
dfnew<-cbind(row,show_data)
change<-colnames(dfnew)
change[1]<-"Variables"
colnames (dfnew)<-change
m1<-gvisTable(dfnew,options=list(showRowNumber=TRUE,width=1220, height=min(400,27*(nrow(show_data)+1)),allowHTML=TRUE,page='disable'))
print(m1,'chart')
```


```{r echo=FALSE, comment=NA, warning=FALSE, message=FALSE,results='asis'}
Rotated_Factors_thres <- Rotated_Factors
Rotated_Factors_thres[abs(Rotated_Factors_thres) < MIN_VALUE]<-NA
colnames(Rotated_Factors_thres)<- colnames(Rotated_Factors)
rownames(Rotated_Factors_thres)<- rownames(Rotated_Factors)

show_data <- Rotated_Factors_thres 
#show_data = show_data[1:min(max_data_report,nrow(show_data)),]
row<-rownames(show_data)
dfnew<-cbind(row,show_data)
change<-colnames(dfnew)
change[1]<-"Variables"
colnames (dfnew)<-change
m1<-gvisTable(dfnew,options=list(showRowNumber=TRUE,width=1220, height=min(400,27*(nrow(show_data)+1)),allowHTML=TRUE,page='disable'))
print(m1,'chart')
```


#### Step 6. Save factor scores for subsequent analyses


<div class="row">
<div class="col-md-6">

```{r echo=FALSE, message=FALSE, prompt=FALSE, results='asis'}
NEW_ProjectData <- round(Rotated_Results$scores[,1:factors_selected,drop=F],2)
colnames(NEW_ProjectData)<-paste("Derived Variable (Factor)",1:ncol(NEW_ProjectData),sep=" ")

if (factors_selected >=2){ 
  
  show_data <- as.data.frame(NEW_ProjectData) 
  show_data = show_data[1:min(max_data_report,nrow(show_data)),]
  row<-rownames(show_data)
  dfnew<-cbind(row,show_data)
  change<-colnames(dfnew)
  change[1]<-"Observation"
  colnames (dfnew)<-change
  m1<-gvisTable(dfnew,options=list(showRowNumber=TRUE,width=1220, height=min(400,27*(nrow(show_data)+1)),allowHTML=TRUE,page='disable'))
  print(m1,'chart')
  
  } else {
    print(xtable(NEW_ProjectData, caption="Only 1 derived variable generated", digits=3), type="html",html.table.attributes="class='table table-striped table-hover table-bordered'", caption.placement="top", comment=FALSE, include.rownames=TRUE, include.colnames=TRUE)
    }
```
</div>
</div>
<br> <br>


```{r Fig2, echo=FALSE, comment=NA, results='asis', message=FALSE, echo=FALSE, fig.align='center', fig=TRUE}

if(ncol(NEW_ProjectData) >= 2) {
  df1  <- cbind(NEW_ProjectData[, 1], NEW_ProjectData[, 2])
  df1  <- as.data.frame(df1)
  sca1 <- gvisScatterChart(df1, options=list(legend="none",
                                             lineWidth=0, pointSize=8, hAxis.title="Derived Variable (Factor) 2",
                                             title="Data Visualization Using the top 2 Derived Attributes (Factors)", vAxis="{title:'Derived Variable (Factor) 1'}",
                                             hAxis="{title:'Derived Variable (Factor) 2'}", width=900, height=800))
  print(sca1,'chart')
  } else {
    df2  <- cbind(NEW_ProjectData[, 1], ProjectData[, 1])
    df2  <- as.data.frame(df2)
    sca2 <- gvisScatterChart(df2, options=list(legend="none",
                                               lineWidth=0, pointSize=12, hAxis.title="Derived Variable (Factor) 1",
                                               title="Only 1 Derived Variable: Using Initial Variable", vAxis="{title:'Derived Variable (Factor) 1'}",
                                               hAxis="{title:'Initial Variable (Factor) 1'}", width=900, height=800))
    print(sca2,'chart')
    }

```

Data Analytics is an iterative process, therefore we also checked different scenarios to return to our original raw data and eventually arrived at the five factors.


**Till then...**


